Demography and selection since maize domestication

### 2015-06-17 manuscript.tex
updated the methods--demographic inference

\subsection{Hard sweeps do not shape maize (or teosinte) diversity}
A mutation that is immediately beneficial and positively selected leaves a classical hard
sweep signature in the genome, whereby genetic diversity surrounding
that mutation is reduced as the haplotype where the mutation first
arose increases in frequency to fixation. The prevalence of such hard
sweeps was evaluated by comparing diversity
surrounding non-synonymous and synonymous substitutions in maize and
teosinte since divergence from \emph{Tripsacum}, roughly \textcolor{red}{ZZZ} years before
present \textcolor{red}{(ref here)}. For both taxa, no difference in diversity
surrounding these classes of substitutions was identified (Figure
\ref{hardSweeps}). Additionally, diversity around maize substitutions
not seen in teosinte was investigated. These sites have the potential
to correspond to recent maize sweeps, occuring after the split from
teosinte. Again, no difference was observed between diversity around
synonymous and nonsynonymous substitutions
\textcolor{red}{(Supplemental figure here)}. Together, these
observations suggest that hard sweeps are not a primary form of
selection for either maize or teosinte.

\begin{figure*}[b]
%\vspace*{.05in}
\centering
\includegraphics[width=.45\textwidth]{FigsAndFiles/plotDiversity_TvM_Folded2_Significance_June}
\hspace{0.05\textwidth} \includegraphics[width=.45\textwidth]{FigsAndFiles/plotDiversity_TvT_Folded2_Significance_June}
\caption{Pairwise diversity surrounding synonymous and nonsynonymous
  substitutions in maize and teosinte. A: Maize diversity; B: Teosinte
diversity.}
\label{hardSweeps}
\end{figure*}

\subsection{Patterns of purifying and background selection are
  influenced by demographic history}
Purifying selection refers to the situation where deleterious
mutations arising in a population are continuously selected against. When this
form of selection is operating, it can serve to reduce genetic
variability at linked neutral sites, a phenomenon called
background selection \cite{charlesworth1993}. Purifying and background selection lead
to lower diversity within genes and other functional sites relative
to neutral regions \textcolor{red}{(ref here)}. We investigated purifying
selection in maize and teosinte by evaluating the average magnitude of reduced
diversity within genes and recovery away from genes in both taxa. When standardized by neutral
levels, which were defined as mean diversity at postions at least 0.01 cM distal
from genes, a stronger reduction of diversity
and slower recovery was observed for teosinte than for maize, implying
that purifying selection has left a more pronounced signature in the
teosinte genome (Figure \ref{purify}). This conflicted with our \emph{a priori} hypothesis;
we expected that strong artificial selection since domestication may
have elevated the intensity of purifying selection for maize. We therefore
conducted a parallel analysis based on singleton diversity. As a class,
singleton alleles depict the most recent patterns of evolution, but
also have the lowest effect on pairwise diversity. Therefore, unlike
pairwise diversity, patterns of singleton
diversity reflect recent patterns of evolution. Since our sample size
for maize (n=23) was larger than teosite (n=13), maize singletons were
analyzed directly as well as downsampled to reflect the sample size of
teosinte. For singletons, our definition of
neutral diversity was modified to be mean diversity at positions at
least 0.02 cM distal from genes. When
evaluating the data in this manner, a very different pattern
emerged. Maize singleton diversity was just at least as low as teosinte singleton
diversity near genes, but recovered more slowly
(Figure \ref{purify}), implying that in the
recent past maize has been at least as influenced by purifying selection as
teosinte. Together, these findings suggest that demographic history has
a strong influence on the effect of purifying selection. Historically,
teosinte has had a larger population size than maize, and only
recently has maize population size overcome that of teosinte. Since
the efficacy of purifying selection scales with population size,
these results likely reflect changes in $N_e$ more than they reflect
underlying changes in selection pressure.

\begin{figure*}[b]
%\vspace*{.05in}
\centering
\includegraphics[width=.45\textwidth]{FigsAndFiles/distanceToGene_WithSignificance_Folded2_manuscript.png} \includegraphics[width=.45\textwidth]{FigsAndFiles/distanceToGene_WithSignificance_Singletons_Downsampled_threeLines_manuscript.png}
\caption{Relative level of diversity versus distance to the nearest
  gene, in maize and teosinte. Two measures of diversity were
  investigated. \textbf{A} displays pairwise
  diversity, which is most influenced by intermediate frequency
  alleles and therefore depicts more ancient evolutionary patterns,
  and \textbf{B} depicts singleton diversity, influenced by rare
  alleles and thus depicting evolutionary patterns in the recent past. \jri{are these done with latest values we used for curve fitting? should also add the singleton vs pi comparison within species as a supplemental figure}}
\label{purify}
\end{figure*}

@@

The most likely model suggested an ancestral for $\theta$ ($4N\mu$) of
0.014734, which is similar to previous estimates
\textcolor{red}{(references here)}. Assuming a mutation rate of $\mu =
3 \times 10^{-8}$ \textcolor{red}{(reference here)}, this suggests an effective
population size of $N_a = 122,783$ individuals, a population split
$T_B = 15,523$ generations in the past, a maize bottleneck
effective populations size of $N_b = 6,455$ individuals (5.26\% of
$N_a$), a modern maize effective population size of $N_m = 366,973$
individuals, $M_{tm} = 1.35$ migrants per generation from teosinte to
maize, and $M_{mt} = 1.72$ migrants per generation from maize to
teosinte (Figure \ref{bottleneck}). We note that a population
split over 15 thousand generations before present precedes estimates
from
archaeological data which suggests maize domestication began
approximately 9,000 years before present \textcolor{red}{(reference
  here)}. This could result from multiple generations per year, or in
may reflect teosinte population structure that was present before
domestication. Also, note that the genetic time of the population
split must precede morphological changes that could be identified
morphologically.   Additionally, because recent expansion is
most evidenced by rare alleles, and since these data provide low
power to detect rare alleles, we expect that the estimate of  $N_m =
366,973$, or $\sim 3N_a$, is likely an underestimate.

We utilized a complementary dataset of 4,808 maize landrace samples collected
from across the Americas \textcolor{red}{(SeeDs reference here)} to
obtain a less downward-biased estimate of the current maize effective
population size. According to a singleton-based estimate of $\theta$
\cite{fu1993}, which was chosen since rapid post-bottleneck expansion
implies that rare alleles will most accurately reflect current maize
demography, we obtained the estimate \textcolor{red}{$N_e =
  1,011,686$}. Although less biased than the previously mentioned
esitmate, we note that ascertainment bias of the
genotyping platform used to generate these data again biases this
estimate downward.

\begin{figure}[b]
%\vspace*{.05in}
\centering
\includegraphics[width=.4\textwidth]{FigsAndFiles/DomesticationModel/domesticationModel.pdf}
\caption{Parameters of domestication as estimated by dadi. Notably,
  the maize effective population size ($N_e$) during the domestication
  bottleneck appears to have consisted of approximately 5.26\%
  the ancestral $N_e$, before recovering two at least 2.98
  times as large as the ancestral $N_e$.}
\label{bottleneck}
\end{figure}

@@

\subsection{Estimating the site frequency spectrum}
To estimate individual  and joint site frequency spectra (SFS) for
maize and teosinte from inbred lines, each inbred individual was treated as representing
a single haplotype from its population. These were
separately computed for genic and intergenic regions, as well as for
the whole genome together. First, genic and intergenic regions were isolated using
the biomaRt package \cite{durinck2009,durinck2005} of R \cite{R2014}. Genic
regions were defined as DNA between the start and stop position of a
gene, while intergenic regions were required to be at least 5kb up- or
down-stream from a gene start/stop. With regions defined, SFS were
estimated with ANGSD \cite{korneliussen2014}. Individual population SFS were
estimated using all positions observed in at least 80\% of the
individuals in the population, and joint SFS were estimated using all
positions observed in at least 80\% of individuals in both
populations. Individuals were
assumed to be fully inbred (-doSaf 2), and subsequently allele frequencies were
divided by two to indicate haplotype frequencies. Quality filters were
employed such that reads with quality score below 30 and bases with
quality score below 20 were discarded (-minMapq 30 and -minQ 20), as were reads that didn't map
uniquely (-uniqueOnly 0). Quality scores around indels were adjusted
as in Samtools (-baq 0). Genotype likelihoods were estimated using the samtools
method (-GL 1). Major and minor alleles were inferred from the data
(-doMaf 1). Because ANGSD cannot calculate a folded joint SFS, the
maize reference genome was used for polarization and then unfolded spectra
were folded using dadi \cite{gutenkunst2009}.

\subsection{Demographic inference}

>>>>

\subsection{Hard sweeps do not shape maize (or teosinte) diversity}
A mutation that is immediately beneficial and positively selected leaves a classical hard
sweep signature in the genome, whereby genetic diversity surrounding
that mutation is reduced as the haplotype where the mutation first
arose increases in frequency to fixation. The prevalence of such hard
sweeps was evaluated by comparing diversity
surrounding non-synonymous and synonymous substitutions in maize and
teosinte since divergence from \emph{Tripsacum}, roughly \textcolor{red}{ZZZ} years before
present \textcolor{red}{(ref here)}. For both taxa, no difference in diversity
surrounding these classes of substitutions was identified (Figure
\ref{fig:hardSweeps}). Additionally, diversity around maize substitutions
not seen in teosinte was investigated. These sites have the potential
to correspond to recent maize sweeps, occuring after the split from
teosinte. Again, no difference was observed between diversity around
synonymous and nonsynonymous substitutions
\textcolor{red}{(Supplemental figure here)}. Together, these
observations suggest that hard sweeps are not a primary form of
selection for either maize or teosinte.

\begin{figure*}[b]
%\vspace*{.05in}
\centering
\includegraphics[width=.45\textwidth]{FigsAndFiles/plotDiversity_TvM_Folded2_Significance_June}
\hspace{0.05\textwidth} \includegraphics[width=.45\textwidth]{FigsAndFiles/plotDiversity_TvT_Folded2_Significance_June}
\caption{Pairwise diversity surrounding synonymous and nonsynonymous
  substitutions in maize and teosinte. A: Maize diversity; B: Teosinte
diversity. \label{fig:hardSweeps}}
\end{figure*}

\subsection{Patterns of purifying and background selection are
  influenced by demographic history}
Purifying selection refers to the situation where deleterious
mutations arising in a population are continuously selected against. When this
form of selection is operating, it can serve to reduce genetic
variability at linked neutral sites, a phenomenon called
background selection \cite{charlesworth1993}. Purifying and background selection lead
to lower diversity within genes and other functional sites relative
to neutral regions \textcolor{red}{(ref here)}. We investigated purifying
selection in maize and teosinte by evaluating the average magnitude of reduced
diversity within genes and recovery away from genes in both taxa. When standardized by neutral
levels, which were defined as mean diversity at postions at least 0.01 cM distal
from genes, a stronger reduction of diversity
and slower recovery was observed for teosinte than for maize, implying
that purifying selection has left a more pronounced signature in the
teosinte genome (Figure \ref{fig:purify}). This conflicted with our \emph{a priori} hypothesis;
we expected that strong artificial selection since domestication may
have elevated the intensity of purifying selection for maize. We therefore
conducted a parallel analysis based on singleton diversity. As a class,
singleton alleles depict the most recent patterns of evolution, but
also have the lowest effect on pairwise diversity. Therefore, unlike
pairwise diversity, patterns of singleton
diversity reflect recent patterns of evolution. Since our sample size
for maize (n=23) was larger than teosite (n=13), maize singletons were
analyzed directly as well as downsampled to reflect the sample size of
teosinte. For singletons, our definition of
neutral diversity was modified to be mean diversity at positions at
least 0.02 cM distal from genes. When
evaluating the data in this manner, a very different pattern
emerged. Maize singleton diversity was just at least as low as teosinte singleton
diversity near genes, but recovered more slowly
(Figure \ref{fig:purify}), implying that in the
recent past maize has been at least as influenced by purifying selection as
teosinte. Together, these findings suggest that demographic history has
a strong influence on the effect of purifying selection. Historically,
teosinte has had a larger population size than maize, and only
recently has maize population size overcome that of teosinte. Since
the efficacy of purifying selection scales with population size,
these results likely reflect changes in $N_e$ more than they reflect
underlying changes in selection pressure.

\begin{figure*}[b]
%\vspace*{.05in}
\centering
\includegraphics[width=.45\textwidth]{FigsAndFiles/distanceToGene_WithSignificance_Folded2_manuscript.png} \includegraphics[width=.45\textwidth]{FigsAndFiles/distanceToGene_WithSignificance_Singletons_Downsampled_threeLines_manuscript.png}
\caption{Relative level of diversity versus distance to the nearest
  gene, in maize and teosinte. Two measures of diversity were
  investigated. \textbf{A} displays pairwise
  diversity, which is most influenced by intermediate frequency
  alleles and therefore depicts more ancient evolutionary patterns,
  and \textbf{B} depicts singleton diversity, influenced by rare
  alleles and thus depicting evolutionary patterns in the recent
  past. \label{fig:purify} \jri{are these done with latest values we used for curve fitting? should also add the singleton vs pi comparison within species as a supplemental figure}}
\end{figure*}

@@

The most likely model suggested an ancestral for $\theta$ ($4N\mu$) of
0.014734, which is similar to previous estimates
\textcolor{red}{(references here)}. Assuming a mutation rate of $\mu =
3 \times 10^{-8}$ \textcolor{red}{(reference here)}, this suggests an effective
population size of $N_a = 122,783$ individuals, a population split
$T_B = 15,523$ generations in the past, a maize bottleneck
effective populations size of $N_b = 6,455$ individuals (5.26\% of
$N_a$), a modern maize effective population size of $N_m = 366,973$
individuals, $M_{tm} = 1.35$ migrants per generation from teosinte to
maize, and $M_{mt} = 1.72$ migrants per generation from maize to
teosinte (Figure \ref{fig:bottleneck}). We note that a population
split over 15 thousand generations before present precedes estimates
from
archaeological data which suggests maize domestication began
approximately 9,000 years before present \textcolor{red}{(reference
  here)}. This could result from multiple generations per year, or in
may reflect teosinte population structure that was present before
domestication. Also, note that the genetic time of the population
split must precede morphological changes that could be identified
morphologically.   Additionally, because recent expansion is
most evidenced by rare alleles, and since these data provide low
power to detect rare alleles, we expect that the estimate of  $N_m =
366,973$, or $\sim 3N_a$, is likely an underestimate.

We utilized a complementary dataset of 4,021 maize landrace individuals collected
from across the Americas \textcolor{red}{(SeeDs reference here)} to
obtain a less downward-biased estimate of the current maize effective
population size. According to a singleton-based estimate of $\theta$
\cite{fu1993}, which was chosen since rapid post-bottleneck expansion
implies that rare alleles will most accurately reflect current maize
demography, we obtained the estimate \textcolor{red}{$N_e =
  992,713.5$}. Although less biased than the previously mentioned
esitmate, we note that ascertainment bias of the
genotyping platform used to generate these data again biases this
estimate downward.

\begin{figure}[b]
%\vspace*{.05in}
\centering
\includegraphics[width=.4\textwidth]{FigsAndFiles/DomesticationModel/domesticationModel.pdf}
\caption{Parameters of domestication as estimated by dadi. Notably,
  the maize effective population size ($N_e$) during the domestication
  bottleneck appears to have consisted of approximately 5.26\%
  the ancestral $N_e$, before recovering two at least 2.98
  times as large as the ancestral $N_e$. \label{fig:bottleneck} }
\end{figure}

@@

\subsection{Estimating the site frequency spectrum}
To estimate individual  and joint site frequency spectra (SFS) for
maize and teosinte from inbred lines, each inbred individual was treated as representing
a single haplotype from its population. SFS were
separately computed for genic and intergenic regions, as well as for
the whole genome together. First, genic and intergenic regions were isolated using
the biomaRt package \cite{durinck2009,durinck2005} of R \cite{R2014}. Genic
regions were defined as DNA between the start and stop position of a
gene, while intergenic regions were required to be at least 5kb up- or
down-stream from a gene start/stop. With regions defined, SFS were
estimated with ANGSD \cite{korneliussen2014}. Individual population SFS were
estimated using all positions observed in at least 80\% of the
individuals in the population, and joint SFS were estimated using all
positions observed in at least 80\% of individuals in both
populations. Individuals were
assumed to be fully inbred (-doSaf 2). Quality filters were
employed such that reads with quality score below 30 and bases with
quality score below 20 were discarded (-minMapq 30 and -minQ 20), as were reads that didn't map
uniquely (-uniqueOnly 0). Quality scores around indels were adjusted
as in Samtools (-baq 0). Genotype likelihoods were estimated using the samtools
method (-GL 1). Major and minor alleles were inferred from the data
(-doMaf 1). Because ANGSD cannot calculate a folded joint SFS, the
maize reference genome was used for polarization and then unfolded spectra
were folded using dadi \cite{gutenkunst2009}.

\subsection{Demographic inference}
We used dadi \cite{gutenkunst2009} to estimate the parameters of maize
domestication based on diffusion approximation of the 2-dimensional
SFS for maize and teosinte. Our first step was to specify a model, as
shown in (Figure \ref{fig:bottleneck}). In words, our model began with an ancestral teosinte
population with effective population size $N_a$ individuals. The
ancestral population split into
two distinct populations, one of maize and the other of teosinte,
$T_b$ generations in the past. The maize population immediately shrank
to size $N_b$, and then grew exponentially to size $N_{maize}$
today. From the split until today, the teosinte effective population
size remained constant. Each year, $M_{mt}$ individuals migrated from
maize to teosinte, and $M_{tm}$ individuals migrated from teosinte to
maize.The free parameters of our model were $N_a, N_b, N_{maize}, T_b,
M_{mt}, \text{and } M_{tm} $. Python code specifying this model is
provided in \textcolor{red}{(SUPPLEMENT)}. We implimented 1,000 dadi
optimizations of this model using the ``dadi.Inference.optimize\_log\_fmin''
optimizer. In dadi optimizations,
initial values for $N_b, N_{maize}, T_b,
M_{mt}, \text{and } M_{tm}$, respectively, were randomly perturbed up
to a factor of two from 0.02, 3, 0.04, 1e-5, and 1e-5. Lower
bounds were set at 1e-7, 1e-7,  0, 1e-10, and
1e-10, respecively, while upper bounds were respectively set at 2,
200, 1, .001, and .001. $N_a$ did not have an initial value, upper, or
lower bounds because it is estimated from equilibrium ancestral theta
($\theta = 4N_a\mu$), which dadi determines automatically regardless
of model. Population sizes were converted from dadi's
units using a mutation rate estimate of 3e-8
\textcolor{red}{(REFERENCE)}. Among the 1,000 dadi runs, parameters
from that with the lowest log-likelihood were taken as estimates of
the parameters of the domestication process.

Recent population expansion, as is observed for maize, most influences
the abundance of rare alleles. However, our demographic parameter
estimation using dadi suffered from a relatively small sample size of
23 maize individuals. Therefore, our approach likely underestimated
current maize effective population size. To address this, we utilized
a complimentary dataset of 4,021 maize landrace individuals collected
from across the Americas \textcolor{red}{(SeeDs reference here)} to
generate an independent estimate of
$N_{maize}$. These individuals were genotyped using GBS
\cite{elshire2011}, a protocol that suffers from a high rate of
missing data \cite{beissinger2013}. To minimiaze biases imposed from
missing data, we used R to isolate only those SNPs that
were observed in at least 1,500 individuals and subsequently projected the
SFS down to a sample of 500 individuals. Then, we utilized the singleton-based estimate of
theta ($\theta = 4N_e\mu = \frac{S}{L}$) to estimate
modern maize $N_e$ \cite{fu1993}, where $S$ is the total number of
singletons and $L$ is the total length of DNA sequenced. Based on the
GBS protocol implemented \cite{elshire2011}, $L$ is equal to the total
number of SNPs in the dataset.
